PA2 - U.S. NOAA Storm Events Impact Analysis (1950 - 2011)
========================================================

## Synopsis
This report explores U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database (1950-2011). We identify events(top 20) which had the most harmful effects on human population and also pin point events(top 20) leading to greatest economic consequences.
Reference data is [storm data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) and we make use of 7 events("EVTYPE", "FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP") to analyze, conclude and present our investigation.
Our final conclusion is "TORNADO" & "EXCESSIVE HEAT" are having the worst impact on human population and "FLOOD", "HURRICANE/TYPHOON" & "TORNADO" are the top events for economical damages(in hundreds of Billion Dollars).


## Data Processing

Setting the echo for code chunks as TRUE globally.
```{r}
library(knitr)
opts_chunk$set(echo=TRUE)
```

### 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Reading the storm data file and storing it in data frame format for futher analysis.
```{r, cache=TRUE}
dat <- bzfile("repdata-data-StormData.csv.bz2")
sd <- read.csv(dat)
dim(sd)
```

Checking out the labels in data
```{r, cache=TRUE}
names(sd)
```

Subsetting data columns for human population impact analysis.
```{r,cache=TRUE}
nsd <- sd[,c(8, 23, 24)]
nsd[,1] <- as.character(nsd[,1])
aggdata <- aggregate(nsd[,2:3], by=list(nsd$EVTYPE), FUN=sum, na.rm=T)
names(aggdata)[1] <- "EVTYPE"

od <- aggdata[order(-aggdata[,2], -aggdata[,3]), ]
head(od, n=10)
```

#### Plot showing the top 20 events responsible for human fatalities 
```{r}
library(ggplot2)
library(scales)

od <- aggdata[order(-aggdata[,2]), ]
names(od)[1] <- "EVTYPE"
topf <- head(od, n=20)
ggplot(data=topf, aes(x=EVTYPE, y=log10(FATALITIES))) + geom_bar(stat = "identity", fill="#FF0000", colour="black") + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Top 20 U.S. NOAA Events causing FATALITIES") + labs(y=expression(log[10](FATALITIES)), x="U.S. NOAA Event Type") 
```

#### Plot showing the top 20 events responsible for human injuries 
```{r}
od <- aggdata[order(-aggdata[,3]), ]
names(od)[1] <- "EVTYPE"
topi <- head(od, n=20)
ggplot(data=topi, aes(x=EVTYPE, y=log10(INJURIES))) + geom_bar(stat = "identity", fill="#FF0000", colour="black") + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Top 20 U.S. NOAA Events causing INJURIES") + labs(y=expression(log[10](INJURIES)), x="U.S. NOAA Event Type") 
```

#### Top Events causing maximum harm with respect to human population health are:
```{r}
intersect(topf[,1], topi[,1])
```

### 2. Across the United States, which types of events have the greatest economic consequences?

```{r} 
##1. subsetting relevant columns for economic impact analysis.
pd <- sd[,c(8,25,26,27,28)]

##2. unique exponents in property damage and crop damage 
unique((pd[,3]))
unique((pd[,5]))
```

```{r}
##3. converting exponents to uppercase for consistency.
pd[,3] <- toupper(pd[,3])
pd[,5] <- toupper(pd[,5])

##4. given a value and exponent - lets calculate the actual value and return
expval <- function(x, exp = "") {
    switch(exp, `-` = x * -1, `?` = x, `+` = x, `1` = x, `2` = x * (10^2), `3` = x * 
        (10^3), `4` = x * (10^4), `5` = x * (10^5), `6` = x * (10^6), `7` = x * 
        (10^7), `8` = x * (10^8), H = x * 100, K = x * 1000, M = x * 1e+06, 
        B = x * 1e+09, x)
}

##5. Aggregating based on event type and exponents.
pad <- aggregate(pd[,c(2,4)], by=list(pd$EVTYPE, pd$PROPDMGEXP, pd$CROPDMGEXP), FUN=sum, na.rm=T)

##6. Remove rows with vaues 0 - they wont be relevant for analysis.
pf = pad[apply(pad[c(4,5)], 1, function(row) any(row != 0 )), ]

##7. Given an event lets calculate the total dollar impact combining property and crop damage.
nr <- nrow(pf)
total <- numeric()
for (i in 1:nr) {
  val <- expval(pf[i,4], pf[i,2])
  val <- val + expval(pf[i,5], pf[i,3])
  total <- append(total, val)
}

length(total)
head(total)

```

Lets check out the interim data frame.
```{r}
pf$TOTALDMG <- total
names(pf)[1] <- "EVTYPE"
head(pf)
```

Aggregating Total Damage based on unique event types.
```{r}
dmg <- aggregate(pf$TOTALDMG, by=list(pf$EVTYPE), FUN=sum, na.rm=T)
names(dmg) <- c("EVTYPE", "TOTALDMG")
head(dmg)
fdmg <- dmg[order(-dmg[,2]), ]
head(fdmg)
```

#### Plot depicting the top 20 events causing the maximum economic damage(in Billion Dollars).
```{r}
tope <- head(fdmg, n=20)
ggplot(data=tope, aes(x=EVTYPE, y=TOTALDMG/1e+9)) + geom_bar(stat = "identity", fill="#FF0000", colour="black") + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Top 20 U.S. NOAA Events causing Economical Consequences") + labs(y="Economic Damage(in Billion Dollars)", x="U.S. NOAA Event Type")
```

## RESULTS
Below we are presenting the final conclusion (events) again from the previous steps of data processing and analysis.
#### Top Events causing maximum harm with respect to human population health(fatalities & injuries combined) are:
```{r}
intersect(topf[,1], topi[,1])
```

#### Top Events causing maximum economic damage (100's of Billion Dollars):
```{r}
head(fdmg, n=10L)
```

Thanks for reading the report - Have a good day ahead!! :)